A Plasma Torus Around a Young Low-Mass Star

### 2025-03-19 paper/ms.tex
update

%\subsection{Introduction}
M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Investment with JWST has been proceeding accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has concurrently become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that are
on average subject to long circumstellar disk lifetimes \cite{Ribas2015},
to large doses of UV radiation \cite{France2013}, and to a high
incidence of flares and coronal mass ejections \cite{Gunther2020}.
However, despite excellent work in these areas, the properties of the
plasma and magnetospheric environments that bathe young, close-in
exoplanets have remained challenging to quantify.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectra of
stars with a wide range of masses have shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.
To date, none of these spectroscopic variables have shown any
photometric anomalies \cite{Bouma2024}, leaving open the issue of
whether they are related to CPVs.
%A circumstantial argument that they might is that CPVs do respond to
%sudden magnetic field changes: the otherwise long-lived eclipse
%features often disappear following stellar flares
%\cite{Stauffer2017,Bouma2024}.

@@

\begin{table*}
\scriptsize
\setlength{\tabcolsep}{2pt}
\centering
\caption{Selected System Parameters for TIC~141146667}
\label{tab:params}
\begin{tabular}{llcc}
\hline \hline
Parameter & Description & Value & Source\\
\hline 
%
$T_{\rm eff}$\dotfill                   & Effective Temperature (K) \hspace{9pt}\dotfill                 & 2972 $\pm$ 40    & 1 \\
%
%$\log{g_{\star}}$\dotfill              & Surface Gravity (cgs)\hspace{9pt}\dotfill                      & YYY              & X \\
%
$R_\star$\dotfill                       & Stellar radius ($R_\odot$)\dotfill                             & 0.42$\pm$0.02    & 1 \\
%
Age                                     & Stellar age range (Myr)\dotfill                                & 35-150  & 2 \\
%
$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.22$\pm$0.02  & 3 \\
%
$\gamma$\dotfill                        & Systemic radial velocity (\kms)\dotfill                        & 0.61 $\pm$ 1.47  & 4 \\
%
Spec. Type\dotfill                      & Spectral Type\dotfill                                          & M5.5Ve           & 4 \\
%
$P_{\rm rot}$\dotfill                   & Rotation period (hr)\dotfill                                   & $3.930\pm 0.001$ & 5 \\
%
$v_{\rm eq}$\dotfill		                & Equatorial velocity \dotfill                                   &  130$\pm$4       & 6 \\
                                        & \hspace{3pt} ($2\pi R_\star/P_{\rm rot}$) (\kms)	             &                      \\
%
$v_{\rm eq}\sin{i_\star}$\dotfill		    & Projected rotational\dotfill                                   &  138$\pm$8       & 4 \\
                                        & \hspace{3pt} velocity (\kms)	                                 &                      \\
%
$i_\star$\dotfill                       & Stellar inclination\dotfill                                    & 	$>$63           & 4 \\
                                        & \hspace{3pt}  2$\sigma$ lower limit (deg)	                     &                      \\
%
$d$\dotfill                             & Distance (pc)\dotfill                                          & $57.54 \pm 0.09$   & 7 \\
%
$R_{\rm c}$\dotfill		                  & Keplerian corotation\dotfill
  & $1.82 \pm 0.10$  & 6 \\
                                        & \hspace{3pt} radius ($R_\star$)	                               &                      \\
%
$a_1$\dotfill                           & Clump 1 orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  2.1-2.7         & 4 \\
$a_2$\dotfill                           & Clump 2 orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  3.9             & 4 \\
%
$\langle$EW$_{\rm H\alpha}$$\rangle$    & Time-averaged H$\alpha$ line core                              &  X.X             & Y \\ 
                                        & \hspace{3pt} equivalent width (\AA)	                           &                      \\
Range(EW$_{\rm H\alpha}$)$_1$           & H$\alpha$ equiv. width range                                   &  X.X             & Y \\ 
                                        & \hspace{3pt} from Clump 1 (\AA)	                               &                      \\
Range(EW$_{\rm H\alpha}$)$_2$           & H$\alpha$ equiv. width range                                   &  X.X             & Y \\ 
                                        & \hspace{3pt} from Clump 2 (\AA)	                               &                      \\
\hline
\end{tabular}
\begin{flushleft}
\footnotesize{ \textsc{NOTE}---
$^*$ Given only $v\sin i$ and $2\pi R_\star/P_{\rm rot}$, $\cos i=0.11^{+0.11}_{-0.08}$.
Provenances are:
1: SED fit \cite{Bouma2024}.
2: Gaia DR3 photometry and lack of lithium (see~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.
%$^6$Cluster isochrone (MIST adopted; PARSEC compared for quoted
%  uncertainty),
4: Keck/HIRES.
5: TESS light curve.
6: Derived quantity.
7: Gaia DR3 geometric \cite{GaiaDR3}.
%  3: HIRES spectra and specmatch(?),
%$^8$Pre-main-sequence CAMD interpolation (Section~\ref{sec:camd}),
}
\end{flushleft}
\vspace{-0.5cm}
\end{table*}

@@

% TEMPLATE: IRAS041
%
\begin{table}
    \centering
    \begin{tabular}{lcr}
    \hline 
    \hline
    Parameter & Host & Source \\
    \hline 
    \multicolumn{3}{c}{Identifiers} \\
    \hline
    TIC & 141146667 & TESS \\
    Gaia & 860453786736413568 & Gaia\ DR3 \\
    %2MASS & J04154278+2909597 & J04154269+2909558 & 2MASS \\
    %ALLWISE & J041542.77+290959.5 & ... & ALLWISE\\
    \hline
    \multicolumn{3}{c}{Astrometry \& Radial Velocity} \\ 
    \hline
    $\alpha_{2000}$ & 11:05:15.09   & Gaia\ DR3 \\
    $\delta_{2000}$ & +59 15 05.57  & Gaia\ DR3 \\
    $\mu_{\alpha}$ (mas yr$^{-1}$ ) & -73.933 $\pm$ 0.022 & Gaia\ DR3 \\
    $\mu_{\delta}$ (mas yr$^{-1}$ ) &  32.262 $\pm$ 0.024 & Gaia\ DR3 \\
    $\pi$ (mas)                     &  17.324 $\pm$ 0.025 & Gaia\ DR3 \\
    RUWE                            &  1.23               & Gaia\ DR3 \\
    RV (\kms)                       & 0.61 $\pm$ 1.47     & HIRES \\
    \hline
    \multicolumn{3}{c}{Photometry} \\
    \hline
    $TESS$ (mag)                    & 13.283 $\pm$ 0.010 & TIC8\     \\
    $G$ (mag)                       & 14.701 $\pm$ 0.002 & Gaia\ DR3 \\
    $G_{\rm BP}$ (mag)              & 16.664 $\pm$ 0.008 & Gaia\ DR3 \\
    $G_{\rm RP}$ (mag)              & 13.398 $\pm$ 0.006 & Gaia\ DR3 \\
    $G_{\rm BP}$-$G_{\rm RP}$ (mag) &  3.276 $\pm$ 0.010 & Gaia\ DR3 \\
    $J$ (mag)                       & 11.401 $\pm$ 0.022 & 2MASS     \\
    $H$ (mag)                       & 10.793 $\pm$ 0.021 & 2MASS     \\
    $K_s$ (mag)                     & 10.473 $\pm$ 0.016 & 2MASS     \\
    $W1$ (mag)                      & 10.276 $\pm$ 0.023 & ALLWISE   \\ % Cutri+2013 II/328/allwise
    $W2$ (mag)                      & 10.070 $\pm$ 0.020 & ALLWISE   \\
    $W3$ (mag)                      &  9.838 $\pm$ 0.045 & ALLWISE   \\
    %$W4$ (mag)                      &  8.392 $\pm$ NOUNC & ALLWISE \\
	  %
    % NOTE: ROSAT is an emitter.  GALEX detection.
    % Ofek+2011 1.4Ghz emitter.  Helfand+2015 ditto.  Bruzewski+2021.
    \hline
    \multicolumn{3}{c}{Physical Properties} \\
    \hline
    $T_{eff}$ (K) & 2972 $\pm$ 40 & \cite{Bouma2024} SED fit\\
    $R_\star$ ($R_{\odot}$) & 0.42 $\pm$ 0.02 & \cite{Bouma2024} SED fit \\
    $L_\star$ ($L_\odot$) & 0.0126 $\pm$ 0.0012 & \cite{Bouma2024} SED fit \\
    $P_{\rm rot}$ (hours) & $3.930\pm 0.001$ & TESS \\ 
    $v_{\rm eq}$ (\kms)  &  130$\pm$4  & Derived \\
    $v_{\rm eq} \sin i_\star$(km s$^{-1}$) & $138 \pm 8$ & HIRES \\
    $i_\star$($^\circ$) & $>$63 & Derived \\
    $A_V$ (mag) & 0 & \cite{Green2019} \\
    $M_\star$ ($M_{\odot}$)  & 0.22 $\pm$ 0.02  & PARSEC \cite{Chen2014}\\
    ${\rm EW}_{\rm Li}$ (m\AA) & $<$107 & HIRES (2$\sigma$)\\
    $t_{\rm CAMD}$ (Myr) & 30-150 &  Gaia DR3 \\
    $t_{\rm Li,emp}$ (Myr) & $>$20 &  HIRES, \cite{Jeffries2023} \\
    $t_{\rm Li,th}$ (Myr) & $>$35 &  HIRES, \cite{Feiden2016} \\
    $t_{\rm adopted}$ (Myr) & 35-150 &  -- \\
    \hline
    \end{tabular}
		\caption{Properties of \starname.  References:
    Gaia DR3 \cite{GaiaDR3}, TESS \cite{Ricker2015},
    TIC8 \cite{Stassun2019}, 2MASS \cite{Skrutskie2006}, ALLWISE
    \cite{Cutri2014}.}
    \label{tab:stparams}
\end{table}

@@

\bibliography{cpvbib.bib} % common bib file
\bibliographystyle{naturemagfixed}

@@

\item[Acknowledgments]
  LGB is grateful to M.~Jardine for insightful conversations,
  A.~Weinberger for clarification of infrared dust spectroscopy,
  B.~Tofflemire for a clarifying line profile analysis, and to both
  A.~Howard and H.~Isaacson for their assistance reducing the HIRES
  spectra.
  LGB was supported by the Carnegie Institution for Science's Carnegie
  Fellowship and the Heising-Simons 51~Pegasi~b Fellowship.
  Funding for the TESS mission is provided by NASA’s
  Science Mission directorate.
  The TESS mission is a product of millions of hours of work by
  thousands of people, and we thank the many TESS team members for
  their efforts to make the mission a continued success.
  The HIRES data were obtained at the Keck Observatory, which exists
  through a similar scale of community effort.
  We recognize the importance that the summit of Maunakea has always had
  within the indigenous Hawaiian community, and we are deeply grateful 
  for the opportunity to conduct observations from this mountain.

>>>>

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Investment with JWST has been proceeding accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has concurrently become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that are
on average subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to large doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite excellent work in these areas,
the properties of the plasma and magnetospheric environments that
bathe young, close-in exoplanets have remained challenging to
quantify.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectra of stars
with a wide range of masses have shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic
field, and the exact geometry of where the plasma can accumulate is
dictated by the field's topology.  For instance, a tilted dipole field
tends to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.  To date, none of these spectroscopic variables have
shown any photometric anomalies \cite{Bouma2024}, leaving open the
issue of whether they are related to CPVs.

@@

\begin{table*}
\scriptsize
\setlength{\tabcolsep}{2pt}
\centering
\caption{Selected System Parameters for TIC~141146667}
\label{tab:params}
\begin{tabular}{llcc}
\hline \hline
Parameter & Description & Value & Source\\
\hline 
%
$T_{\rm eff}$\dotfill                   & Effective Temperature (K) \hspace{9pt}\dotfill                 & 2972 $\pm$ 40    & 1 \\
%
$R_\star$\dotfill                       & Stellar radius ($R_\odot$)\dotfill                             & 0.42$\pm$0.02    & 1 \\
%
Age                                     & Stellar age range (Myr)\dotfill                                & 35-150           & 2 \\
%
$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.22$\pm$0.02    & 3 \\
%
$\gamma$\dotfill                        & Systemic radial velocity (\kms)\dotfill                        & 0.61 $\pm$ 1.47  & 4 \\
%
Spec. Type\dotfill                      & Spectral Type\dotfill                                          & M5.5Ve           & 4 \\
%
$P_{\rm rot}$\dotfill                   & Rotation period (hr)\dotfill                                   & $3.930\pm 0.001$ & 5 \\
%
$v_{\rm eq}$\dotfill		                & Equatorial velocity \dotfill                                   &  130$\pm$4       & 6 \\
                                        & \hspace{3pt} ($2\pi R_\star/P_{\rm rot}$) (\kms)	             &                      \\
%
$v_{\rm eq}\sin{i_\star}$\dotfill		    & Projected rotational\dotfill                                   &  138$\pm$8       & 4 \\
                                        & \hspace{3pt} velocity (\kms)	                                 &                      \\
%
$i_\star$\dotfill                       & Stellar inclination\dotfill                                    & 	$>$63           & 4 \\
                                        & \hspace{3pt}  2$\sigma$ lower limit (deg)	                     &                      \\
%
$d$\dotfill                             & Distance (pc)\dotfill                                          & $57.54 \pm 0.09$ & 7 \\
%
$R_{\rm c}$\dotfill		                  & Keplerian corotation\dotfill                                   & $1.82 \pm 0.10$  & 6 \\
                                        & \hspace{3pt} radius ($R_\star$)	                               &                      \\
%
$a_1$\dotfill                           & Clump 1 orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  2.1-2.7         & 4 \\
$a_2$\dotfill                           & Clump 2 orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  3.9             & 4 \\
%
$\langle$EW$_{\rm H\alpha}$$\rangle$    & Time-averaged H$\alpha$ line core                              &  X.X             & Y \\ 
                                        & \hspace{3pt} equivalent width (\AA)	                           &                      \\
Range(EW$_{\rm H\alpha}$)$_1$           & H$\alpha$ equiv. width range                                   &  X.X             & Y \\ 
                                        & \hspace{3pt} from Clump 1 (\AA)	                               &                      \\
Range(EW$_{\rm H\alpha}$)$_2$           & H$\alpha$ equiv. width range                                   &  X.X             & Y \\ 
                                        & \hspace{3pt} from Clump 2 (\AA)	                               &                      \\
\hline
\end{tabular}
\begin{flushleft}
\footnotesize{ \textsc{NOTE}---
$^*$ Given only $v\sin i$ and $2\pi R_\star/P_{\rm rot}$, $\cos i=0.11^{+0.11}_{-0.08}$.
Provenances are:
1: SED fit \cite{Bouma2024}.
2: Pre-main sequence Gaia DR3 photometry, but lacks of lithium (see~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.
4: Keck/HIRES.
5: TESS light curve.
6: Derived quantity.
7: Gaia DR3 geometric \cite{GaiaDR3}.
}
\end{flushleft}
\vspace{-0.5cm}
\end{table*}

@@

\begin{table}
    \centering
    \begin{tabular}{lcr}
    \hline 
    \hline
    Parameter & Host & Source \\
    \hline 
    \multicolumn{3}{c}{Identifiers} \\
    \hline
    TIC & 141146667 & TESS \\
    Gaia & 860453786736413568 & Gaia\ DR3 \\
    \hline
    \multicolumn{3}{c}{Astrometry \& Radial Velocity} \\ 
    \hline
    $\alpha_{2000}$ & 11:05:15.09   & Gaia\ DR3 \\
    $\delta_{2000}$ & +59 15 05.57  & Gaia\ DR3 \\
    $\mu_{\alpha}$ (mas yr$^{-1}$ ) & -73.933 $\pm$ 0.022 & Gaia\ DR3 \\
    $\mu_{\delta}$ (mas yr$^{-1}$ ) &  32.262 $\pm$ 0.024 & Gaia\ DR3 \\
    $\pi$ (mas)                     &  17.324 $\pm$ 0.025 & Gaia\ DR3 \\
    RUWE                            &  1.23               & Gaia\ DR3 \\
    RV (\kms)                       & 0.61 $\pm$ 1.47     & HIRES \\
    \hline
    \multicolumn{3}{c}{Photometry} \\
    \hline
    $TESS$ (mag)                    & 13.283 $\pm$ 0.010 & TIC8\     \\
    $G$ (mag)                       & 14.701 $\pm$ 0.002 & Gaia\ DR3 \\
    $G_{\rm BP}$ (mag)              & 16.664 $\pm$ 0.008 & Gaia\ DR3 \\
    $G_{\rm RP}$ (mag)              & 13.398 $\pm$ 0.006 & Gaia\ DR3 \\
    $G_{\rm BP}$-$G_{\rm RP}$ (mag) &  3.276 $\pm$ 0.010 & Gaia\ DR3 \\
    $J$ (mag)                       & 11.401 $\pm$ 0.022 & 2MASS     \\
    $H$ (mag)                       & 10.793 $\pm$ 0.021 & 2MASS     \\
    $K_s$ (mag)                     & 10.473 $\pm$ 0.016 & 2MASS     \\
    $W1$ (mag)                      & 10.276 $\pm$ 0.023 & ALLWISE   \\ % Cutri+2013 II/328/allwise
    $W2$ (mag)                      & 10.070 $\pm$ 0.020 & ALLWISE   \\
    $W3$ (mag)                      &  9.838 $\pm$ 0.045 & ALLWISE   \\
    %$W4$ (mag)                      &  8.392 $\pm$ NOUNC & ALLWISE \\
	  %
    % NOTE: ROSAT is an emitter.  GALEX detection.
    % Ofek+2011 1.4Ghz emitter.  Helfand+2015 ditto.  Bruzewski+2021.
    \hline
    \multicolumn{3}{c}{Physical Properties} \\
    \hline
    $T_{eff}$ (K) & 2972 $\pm$ 40 & \cite{Bouma2024} SED fit\\
    $R_\star$ ($R_{\odot}$) & 0.42 $\pm$ 0.02 & \cite{Bouma2024} SED fit \\
    $L_\star$ ($L_\odot$) & 0.0126 $\pm$ 0.0012 & \cite{Bouma2024} SED fit \\
    $P_{\rm rot}$ (hours) & $3.930\pm 0.001$ & TESS \\ 
    $v_{\rm eq}$ (\kms)  &  130$\pm$4  & Derived \\
    $v_{\rm eq} \sin i_\star$(km s$^{-1}$) & $138 \pm 8$ & HIRES \\
    $i_\star$($^\circ$) & $>$63 & Derived \\
    $A_V$ (mag) & 0 & \cite{Green2019} \\
    $M_\star$ ($M_{\odot}$)  & 0.22 $\pm$ 0.02  & PARSEC \cite{Chen2014}\\
    ${\rm EW}_{\rm Li}$ (m\AA) & $<$107 & HIRES (2$\sigma$)\\
    $t_{\rm CAMD}$ (Myr) & 30-150 &  Gaia DR3 \\
    $t_{\rm Li,emp}$ (Myr) & $>$20 &  HIRES, \cite{Jeffries2023} \\
    $t_{\rm Li,th}$ (Myr) & $>$35 &  HIRES, \cite{Feiden2016} \\
    $t_{\rm adopted}$ (Myr) & 35-150 &  -- \\
    \hline
    \end{tabular}
		\caption{Properties of \starname.  References:
    Gaia DR3 \cite{GaiaDR3}, TESS \cite{Ricker2015},
    TIC8 \cite{Stassun2019}, 2MASS \cite{Skrutskie2006}, ALLWISE
    \cite{Cutri2014}.}
    \label{tab:stparams}
\end{table}

@@

\bibliography{cpvbib.bib}
\bibliographystyle{naturemagfixed}

@@

\item[Acknowledgments]
  LGB is grateful to M.~Jardine, A.~Weinberger, and B.~Tofflemire for
  insightful conversations, and to A.~Howard and H.~Isaacson for their
  assistance reducing the HIRES spectra.
  LGB was supported by the Carnegie Institution for Science's Carnegie
  Fellowship and the Heising-Simons 51~Pegasi~b Fellowship.
  %
  Funding for the TESS mission is provided by NASA’s Science Mission
  directorate.
  The TESS mission is a product of millions of hours of work by
  thousands of people; we thank the TESS team for their efforts to
  make the mission a continued success.
  The HIRES data were obtained at the Keck Observatory, which exists
  through a similar scale of community effort.
  We recognize the importance that the summit of Maunakea has always had
  within the indigenous Hawaiian community, and we are deeply grateful 
  for the opportunity to conduct observations from this mountain.
